---
title: "HP_demo"
author: "Joe Boyd"
date: "2024-04-17"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteIndexEntry{Overview and Use Cases}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Objective

To demonstrate the general use of ChIPtsne2 in making common ChIP-seq plot types : heatmaps, line plots, etc.

# Set up

## Libraries

```{r libs}
library(chiptsne2)
library(GenomicRanges)
#I'm leaving seqsetvis unloaded
#all seqsetvis functions will be  called via seqsetvis::
```

## Data

To use ChIPtsne2, we need 2 things : 

1.  A source of enrichment signals (bam files or bigwig files) to fetch profiles from.
2.  A set of feature regions (GRanges) at which to analyze signal profiles.

### Signal

Since we have an easy way to load a data.frame connecting bam files to metadata, we'll use that.

```{r data signal}
bam_cfg_dt = IKdata::chipseq.setup_bam_files()
DT::datatable(
    bam_cfg_dt[, .(cell, treatment, mark, rep, round, name)], 
    options = list(scrollX = TRUE)
)
```

While not strictly necessary, creating a `FetchConfig` object

First we'll select for the mark of interest and prune extra attributes.

```{r signal FetchConfig}
bam_cfg_dt.sel = bam_cfg_dt[mark == "ERG", .(file, cell, treatment, mark, rep, round, name, mapped_reads)]
fetch_cfg = FetchConfig(bam_cfg_dt.sel, 
                        read_mode = "bam_SE", 
                        view_size = 2e3, 
                        window_size = 50)
```

That should look a lot like making a ssvQC::QcConfigSignal object if you've done that. In fact, if you've used ssvQC::QcConfigSignal.save_config you can use chiptsne2::FetchConfig.load_config to load it into chiptsne2.

### Regions

We need a single GRanges object to query signal from. This will be created by overlapping the individual peak calls.

Like with the bam files we have an easy way to load a data.frame connecting narrowPeak files to metadata.

```{r data regions}
np_cfg_dt = IKdata::chipseq.setup_peak_files()
DT::datatable(
    np_cfg_dt[, .(cell, treatment, mark, rep, round, name)], 
    options = list(scrollX = TRUE)
)
```

#### overlap

We don't need to look at every ChIP-seq at once, so let's select one mark to demonstrate.

We also need to reduce multiple peak calls to a single GRanges. We'll use *seqsetvis* for this.

```{r load peak overlap}
np_cfg_dt.sel = np_cfg_dt[mark == "ERG",]
np_grs = seqsetvis::easyLoad_narrowPeak(np_cfg_dt.sel$file, file_names = np_cfg_dt.sel$name)
olap_gr = seqsetvis::ssvOverlapIntervalSets(np_grs)
```

#### plots

bar plot

```{r features bars}        
seqsetvis::ssvFeatureBars(olap_gr)
```

binary heatmap

```{r features binary}
seqsetvis::ssvFeatureBinaryHeatmap(olap_gr)
```

Euler plot

```{r features euler}
seqsetvis::ssvFeatureEuler(olap_gr)
```

Upset plot

```{r features upset}
seqsetvis::ssvFeatureUpset(olap_gr)
```

#### final query

```{r final query}
query_gr = olap_gr
#subset for regions where replicates agree
query_gr = subset(query_gr, 
                  (MXP5_ERG_GFP_rep1 & MXP5_ERG_GFP_rep2) | 
                      (MXP5_ERG_IK_rep1 & MXP5_ERG_IK_rep2)
)
query_gr
```

This yields about 7k regions, which is a reasonable number.

# ChIPtsne2 object

After setting up our query regions and config for fetching signal we can move onto the main part of ChIPtsne2.

Since we're using a FetchConfig, chiptsne2 will fetch the signal profiles as part of creating this object.

If you already have profiles from a seqsetvis::ssvFetch* call, you can use 

```{r create chiptsne2}
ct2 = ChIPtsne2.from_FetchConfig(fetch_config = fetch_cfg, query_gr = query_gr)
# setting the seed ensures any probabilistic functions (k-means, tSNE) return identical results each time
ct2 = setSeed(ct2, seed = 0)
```

## heatmap and lineplot

With this object we can go right into a heatmap.

```{r}
plotSignalHeatmap(ct2)
```
Or a lineplot.

```{r}
plotSignalLinePlot(ct2)
```

## metadata and grouping

It's important to understand that we haven't added clustering information yet to this ChIptsne2 object. All it knows it what was in the original query.

Let's look at the information that is already present.

```{r}
rowRanges(ct2)
```

We can bring any of these into our plots like so:

```{r}
plotSignalHeatmap(ct2, group_VARS = "region_MXP5_ERG_GFP_rep1")
plotSignalLinePlot(ct2, group_VAR = "region_MXP5_ERG_GFP_rep1")
```

And even more than one at once.

```{r}
plotSignalHeatmap(ct2, group_VARS = 
                      c("region_MXP5_ERG_GFP_rep1", "region_MXP5_ERG_GFP_rep2", 
                        "region_MXP5_ERG_IK_rep1", "region_MXP5_ERG_IK_rep2"), 
                  annotation_colors = c("FALSE" = "gray", "TRUE" = "black"),
                  color_key_strategy = "all")
plotSignalLinePlot(ct2, group_VAR = "region_MXP5_ERG_GFP_rep1", color_VAR = "region_MXP5_ERG_IK_rep1")
```

We can add region metadata like so.

```{r}
ct2 = mutateRegions(ct2, GFP = region_MXP5_ERG_GFP_rep1 & region_MXP5_ERG_GFP_rep2)
ct2 = mutateRegions(ct2, IK = region_MXP5_ERG_IK_rep1 & region_MXP5_ERG_IK_rep2)
ct2 = groupRegionsByMembershipTable(ct2, membership = rowData(ct2)[, c("GFP", "IK")])
rowData(ct2)
```

```{r}
plotSignalHeatmap(ct2, group_VARS = 
                      c("GFP", "IK", "membership_id"), 
                  annotation_colors = c("FALSE" = "gray", "TRUE" = "black")
)
plotSignalLinePlot(ct2, group_VAR = "GFP", color_VAR = "IK")
```

## clustering

As far as chiptsne2 is concerned, clustering is just a special kind of grouping based on signal profiles.

This is calling seqsetvis::ssvSignalClustering, i.e. k-means. So you should expect similar results.

```{r kmeans clustering}
ct2 = groupRegionsBySignalCluster(ct2, group_VAR = "cluster_id", n_clusters = 6)
#this has added cluster_id to region data
rowData(ct2)
```

Now we can use our new group_VAR.

```{r}
plotSignalHeatmap(ct2, 
                  group_VARS = c("cluster_id")
)
plotSignalLinePlot(ct2, group_VAR = "cluster_id")
```

Uh oh! Looks like we have a bad cluster there

```{r}
ct2.sub = subsetRegions(ct2, cluster_id != 1)
plotSignalHeatmap(ct2.sub, 
                  group_VARS = c("cluster_id")
)
rowData(ct2)
rowData(ct2.sub)$cluster_id = droplevels(rowData(ct2.sub)$cluster_id)
all(rownames(rowData(ct2.sub)) == rownames(ct2.sub))
debug(plotSignalLinePlot)
plotSignalLinePlot(ct2.sub, 
                   group_VAR = "cluster_id")
```
